From Causal Discovery to Intervention Simulation: Modeling Precariousness and Depression in the HELIUS Study

### 2025-08-25 draft/draft_v3.qmd
add draft_v3 (nature format)

affiliations:
  - id: 1
    name: \textit{Computational Science Lab, Informatics Institute, University of Amsterdam, PO Box 94323, Amsterdam, 1090GH, the Netherlands}
  - id: 2
    name: \textit{Department of Public and Occupational Health, Amsterdam Public Health Research Institute, Amsterdam UMC, University of Amsterdam, Amsterdam, the Netherland}
  - id: 3
    name: \textit{Institute for Advanced Study, University of Amsterdam, Oude Turfmarkt 147, Amsterdam, 1012GC, the Netherland}   
format: 
  # docx:
  #   number-sections: true
  #   appendix: true
  pdf:
    pdf-engine: pdflatex
    fontfamily: palatino
    #toc: true
    number-sections: true
    appendix: true
    colorlinks: true
    keep-tex: true
    geometry: 
      #- showframe
      - top=40mm
      - left=45mm
      - right=45mm
      - bottom=40mm
    template-partials:
      - title.tex
    include-in-header: header.tex
    crossref:
      appendix-title: "App."
      appendix-delim: ":"
date: today
execute:
  echo: false
  warning: false
bibliography: references_draft.bib
nocite: |
  @*
csl: apa.csl
abstract: |
  \noindent This study investigates how the internal structure of depression–precariousness dynamics shapes responsiveness to external support. Using cross-sectional data from the HELIUS study, we first apply cycle-capable causal discovery algorithms to estimate directional dependencies between financial stress, domain-specific precariousness, and depressive symptoms. Our multi-resolution analysis—spanning both aggregate and symptom-level constructs—highlights financial stress as a plausible upstream driver of depression, with symptoms such as sleep disturbance and depressed mood emerging as initiators of broader social vulnerability.
  To translate these structural insights into dynamic predictions, we construct a family of analytically calibrated linear stochastic models. Simulations reveal that variations in internal structure can substantially alter system responses to financial stress reduction. Some models show rapid but transient improvements in depression, while others adjust more slowly yet sustain gains over time. Precariousness tends to respond more immediately across all configurations, though its trajectory also shifts subtly with structural variation.
  Together, these findings demonstrate that intervention outcomes depend not only on the magnitude of external input but also on latent properties of the system. More broadly, this study illustrates how combining causal discovery with dynamical modeling supports principled, simulation-based exploration of intervention scenarios—grounded in data, yet able to probe counterfactual structure. This approach offers a scalable framework for diagnosing hidden system constraints and informing targeted public health strategies in complex psychosocial domains.
---

@@

This study aims to investigate the causal mechanisms linking precariousness and depression and how those mechanisms shape system responses to external intervention. Specifically, we ask: *How does the internal configuration of the precariousness–depression relationship affect the outcomes of stress-reducing interventions?*
To answer this, we combine two approaches. First, we use cycle-capable causal discovery algorithms to infer directional relationships among domain-specific precariousness factors and depressive symptoms.
By examining both aggregated measures (depression sum score and a composite precariousness index) and disaggregated variables (individual depressive symptoms and specific life-domain indicators), we capture broad patterns of relationship as well as fine-grained causal pathways.
Second, we translate these insights into a computational dynamical model that simulates how depression and precariousness co-evolve under varying levels of external stress. 
This model enables us to probe how system properties—such as feedback strength and stochastic noise—govern responsiveness to intervention. 
When the effects of depression symptoms onto the social precarity subsystem are weaker than the reverse path, there is higher sensitivity of depression to intervention, reflecting both a faster reduction post financial intervention but also quicker relapse once financial stresses are reestablished.
By integrating causal discovery with dynamic simulation, we demonstrate that intervention strategies can be tailored to the underlying architecture of vulnerability.

<!-- Mental health problems are rising globally, and populations facing chronic stress and socioeconomic instability bear a disproportionate share of this burden. Despite growing recognition of these challenges, designing effective policies remains difficult—largely because mental health outcomes emerge from complex, interacting factors that do not operate in isolation. To inform better interventions, we must move beyond identifying which factors are associated with mental health problems and instead uncover how those factors influence one another, the underlying causal mechanisms. -->

<!-- Precarious life conditions—such as unstable housing, insecure employment, and financial hardship—are now well established as key correlates of depression and other mental health outcomes. However, the causal dynamics linking these domains remain poorly understood. In particular, it is unclear how different forms of precariousness and depressive symptoms influence one another, and whether their interplay creates self-reinforcing loops that stabilize poor mental health and social adversity over time.  -->

<!-- This study aims to investigate the causal mechanisms linking precariousness and depression, and to examine how those mechanisms shape the system’s response to external intervention. Specifically, we ask: How do different configurations of the precariousness–depression relationship affect the outcomes of interventions aimed at reducing external stress? -->

<!-- To address this, we combine causal discovery and mechanistic modeling. We apply cycle-capable causal discovery algorithms to infer directional relationships among financial stress, precariousness indicators, and depressive symptoms. Our analysis includes both aggregate scores (e.g., PHQ-9 total, precariousness composite index) and disaggregated variables (individual symptoms and domain-specific precariousness), allowing us to capture both broad associations and fine-grained pathways. -->

<!-- Building on the discovered causal patterns, we simulate a set of scenarios using a computational model that captures the co-evolution of depression and precariousness under varying levels of financial stress. This framework allows us to examine how internal system properties—particularly feedback strength and stochastic noise—influence responsiveness to intervention. In some regimes, the system exhibits bistability, resisting change unless pushed past a critical threshold. In others, it transitions gradually, reflecting unistable dynamics. Through this integrative approach, we aim to show how the internal architecture of the depression–precariousness system fundamentally shapes its susceptibility to change—highlighting the importance of structural understanding in designing effective, equitable mental health interventions. -->

# Methods

<!-- To ensure robust estimation and guard against sensitivity to parameter settings, we implemented a systematic grid of analysis conditions: -->

<!-- * Significance levels: $\alpha$ = 0.01 and 0.05 -->
<!-- * Bootstrap thresholds: 0.5, 0.6, 0.7, 0.8 -->
<!-- * CI tests: Gaussian and RCoT -->
<!-- * Algorithms: FCI, CCI, and PC -->

<!-- Each combination was bootstrapped 100 times for sum-score level analyses and 30 times for symptom-level analyses, yielding a total of 1,600 graphs per algorithm for the full analysis. For symptom networks, we fixed the skeleton structure using a consensus graph estimated across PC, FCI, and CCI ($\alpha$ = 0.01, RCoT). This constraint improves computational efficiency without limiting the estimation of directional edges between symptoms and precariousness variables. -->

<!-- To summarize the results, we adopted a two-level aggregation strategy. First, within each analysis condition, we recorded the most frequently observed edge-endpoint type (e.g., directed, undirected, uncertain) from the bootstraps. Then, across all conditions, we tallied the most frequent endpoint type per edge. In cases of ties, we assigned dashed edges to reflect uncertainty. For transparency, we accompany each graph with a table showing the relative frequencies of each edge symbol. -->

<!-- This procedure yields a final graph for each algorithm and analysis type—offering both robustness and interpretability. See @fig-workflow for an overview of the causal discovery pipeline. -->
<!-- For detailed explanations of the algorithms, edge interpretations, and graph types (e.g., PAG, MAAG, CPDAG), as well as algorithmic assumptions and limitations, see the [Appendix](#sec-appendix). -->

<!-- To uncover directional relationships between precariousness and depression, we used constraint-based causal discovery algorithms capable of detecting feedback loops and latent confounding. Specifically, we applied the *Fast Causal Inference* (FCI) and *Cyclic Causal Inference* (CCI) algorithms, both of which support cyclic and non-acyclic structures [@mooij2020constraint; @strobl2019]. Additionally, we include the *PC* algorithm as a reference, given its simplicity and prominence as one of the most widely known causal discovery methods [@spirtes2001causation]. -->

<!-- One practical challenge in applying these algorithms to the HELIUS dataset is that most variables do not follow a Gaussian distribution, and the relationships among variables are unlikely to be strictly linear. To address this, we supplement the commonly used Gaussian conditional independence (CI) test, which relies on partial correlations, with a non-parametric CI test based on reproducing kernel methods [@zhang2012kernel]. -->
<!-- While kernel-based conditional independence tests (KCIT) are theoretically effective in capturing non-linear dependencies and subtle interactions in data with non-Gaussian noise, they are computationally demanding for large datasets like HELIUS. Their quadratic scaling with sample size is primarily due to the need to invert large kernel matrices, making them impractical for large-scale analysis [@rahimi2007random]. -->
<!-- To mitigate this issue, we use the Randomized Conditional Correlation Test (RCoT), which approximates kernel methods using random Fourier features. This approach reduces computational complexity from quadratic to linear scaling with sample size, significantly lowering computational costs while maintaining robustness [@strobl2019approximate]. For a more detailed explanation of RCoT, refer to @sec-rcot. -->

<!-- We analyze the causal structure using three complementary approaches: (1) Examining the relationship between the five individual precariousness factors and the PHQ sum score, which represents overall depression severity. -->
<!-- (2) Investigating the relationship between individual depressive symptoms and individual precariousness factors. (3) Assessing the relationship between individual depressive symptoms and overall precariousness (the sum of all precariousness factors). -->
<!-- Using the PHQ sum score as a measure of depression simplifies the analysis by reducing dimensionality, improving computational efficiency, and providing a broad, interpretable perspective on how precariousness factors relate to overall depression severity. However, this aggregated approach may overlook granular symptom-level relationships. -->

<!-- By zooming in on individual symptoms, we capture a more detailed picture of how specific depressive symptoms interact with different precariousness factors. This approach allows us to map individual causal pathways, identifying which symptoms are particularly sensitive to certain precarious conditions and vice versa. However, this level of detail introduces methodological challenges, including non-standard distributions of symptom variables and the high-dimensional nature of the analysis, which can reduce statistical power and significantly increase computational demands—especially when using non-parametric conditional independence (CI) tests. -->

<!-- Finally, analyzing individual symptoms in relation to overall precariousness provides an integrative perspective, revealing how depression symptoms collectively relate to precarious conditions. This approach complements the fully disaggregated analysis by aggregating precariousness factors, potentially uncovering relationships that may be too weak to detect when examined separately. This is particularly relevant when a symptom is associated with multiple precariousness factors, but these associations are not strong enough to emerge in an individual-level analysis. -->
<!-- By integrating these three approaches — ranging from partially aggregated to fully disaggregated — we aim to achieve a balance between capturing detailed granularity and gaining a broader understanding of the relationship between precariousness and depression.  -->

<!-- Furthermore, to address sensitivity to parameter values and enhance the robustness of our results, we implement multiple settings combined with bootstrapping. For each condition, we generate 100 bootstrap samples, estimate causal graphs, and retain only the edge-endpoints that exceed predefined thresholds. If no edge-endpoint type surpasses the threshold, we assign a circle ($\circ$) to indicate uncertainty in the edge direction. -->

<!-- The analyses are conducted under the following conditions: -->

<!-- -   Significance levels ($\alpha$): 0.01 and 0.05 -->
<!-- -   Thresholds: 0.5, 0.6, 0.7, and 0.8 -->
<!-- -   CI test: Gaussian CI test, RCoT -->
<!-- -   Algorithms: FCI, CCI, and PC -->

<!-- This setup yields 16 combinations (2 significance levels × 4 thresholds × 2 CI tests), applied across three algorithms. Each combination is repeated for 100 bootstrap samples, yielding a total of 1,600 graphs per algorithm.  -->
<!-- For analyses involving individual symptom variables, we simplify the setup by focusing on thresholds of 0.6 and 0.7 and reducing the number of bootstrap samples to 30, easing computational demands. -->
<!-- To further enhance efficiency, we fix the skeleton of the symptom network using a common structure derived from PC, FCI, and CCI algorithms ($\alpha$ = 0.01, RCoT CI test). This step constrains which edges can be present among symptoms but does not restrict the estimation of causal directions within the symptom network. Additionally, the entire causal structure between symptoms and precariousness factors remains fully estimable, ensuring that key relationships are still inferred. -->
<!-- Beyond its computational benefits — particularly in reducing the time required for skeleton estimation — this fixed structure also aligns well with commonly reported symptom network structures in the literature [@park2025individual], ensuring consistency with existing findings. -->

<!-- To obtain a final, stable summary of causal relationships, we identify the most frequently occurring edge-endpoints across different experimental setups. First, within each condition, we determine the most frequent edge-endpoint from all bootstrap samples. Then, across different conditions, we aggregate these selected symbols and determine the overall most common edge-endpoint for each graph entry. -->

<!-- Since we have an even number of experimental setups, ties may occasionally occur. In such cases, we explicitly denote them in the resulting graph using dashed lines. To further help with the interpretation and understanding of the resulting graph, we provide a corresponding table for each graph, showing the proportion of each edge-endpoint type to clarify how dominant a given symbol is in each case. -->

<!-- By double summarizing the results in this way, we generate one final graph per algorithm for each analysis approach: the partially aggregated analysis, which uses either the sum of precariousness factors or the sum of depression factors to capture overall trends, and the fully disaggregated analysis, which examines individual precariousness and depression factors separately. -->
<!-- This approach ensures that only stable and consistent edges — those that persist across different parameter settings — are retained, leading to a robust and reliable understanding of the relationships between precariousness factors and depression. See @fig-workflow for an overview of the analysis workflow. -->

```{r}
#| label: fig-concept
#| fig-cap: "High-level schematic of the dynamic model structure used in simulation. Arrows represent influence coefficients $\\alpha$, matching the terms in the system of stochastic differential equations. $D_i(t)$ and $P_i(t)$ denote depressive symptoms and social precariousness for individual $i$, with $S_i$ representing static financial stress."
#| fig-align: center
#| out-width: 40%

@@

@@

<!-- In  graph  (a),  the  key  pathways  suggest  that  employment  precariousness  (P.emp) and social precariousness (P.soc) do not cause depression (PHQsum).P.soc does not cause recent relational stress (S.rel) or financial stress (S.fin), and these stressors are likely related by a latent confounder.   Additionally, P.emp and S.fin are identified as non-causes of housing precariousness (P.hou). Both FCI and CCI detect a dependency between P.emp and S.fin, but they disagree on the direction of the relationship, leaving it unclear whether S.fin causes P.emp or whether a latent confounder mediates their relationship. Similar ambiguities are found in the connections between S.rel and PHQsum and between S.fin and PHQsum -->

<!-- Both FCI and CCI detect a dependency between *P.emp* and *S.fin*, but they disagree on the direction of the relationship, leaving it unclear whether *S.fin* causes *P.emp* or whether a latent confounder mediates their relationship. Similar ambiguities are found in the connections between *S.rel* and *PHQsum* and between *S.fin* and *PHQsum*. -->

<!-- Graph (b), derived solely from the nonparametric RCoT method. Both graphs consistently identify that *P.emp* is not a cause of *P.hou* or *PHQsum*, and that *P.soc* does not cause *PHQsum* or *S.fin*.  -->
<!-- The relationship between stressors and depression remains unresolved between the two algorithms. However, the graph derived from RCoT provides greater confidence that depression is not the cause of *S.fin* and suggests a stronger likelihood that the financial stress may contribute to depression or that their relationship is mediated by a latent confounder. -->
<!-- The role of relational stress has become less pronounced, as the edge between *P.soc* and *S.rel* is omitted, and the relationship between *S.rel* and *S.fin* is now inconsistent between the algorithms. -->

<!-- Overall, both graphs together highlight the potential roles of stressors like *S.fin* and *S.rel*, which are closely tied to depression, either causally or via latent confounders. Employment and social precariousness may also be influenced by depression, either directly or through latent variables. Lastly, housing precariousness does not directly impact depression but remains connected to employment precariousness and financial stress, likely mediated by direct or latent pathways. -->

@@

```{r}
#| label: fig-intervention
#| fig-align: center
#| out-width: 80%
#| fig-cap: >
#|   System response to external intervention across variations in internal structure.
#|   ***Panel A**: Mean levels of depression ($D$) and precariousness ($P$) at time steps 100, 300, and 500, across varying values of $\alpha_{DP}$ and intervention strengths. The x-axis represents the strength of external support, operationalized as the proportional reduction of financial stress ($S$) toward its minimum observed value. Each line corresponds to a different model defined by a specific $\alpha_{DP}$ value, color-coded from low (purple) to high (yellow). Models with lower $\alpha_{DP}$ exhibit faster early reductions in $D$, while higher $\alpha_{DP}$ leads to more immediate declines in $P$.
#|   
#|   **Panel B**: Full time trajectories for two models—one with no coupling ($\alpha_{DP} = 0$) and one with strong coupling ($\alpha_{DP} = 0.65$)—illustrating system response and recovery following intervention onset (step 300) and withdrawal (step 1500).
#|   Depression responds more gradually and persistently in the high $\alpha_{DP}$ model, whereas precariousness adjusts more quickly.*

@@

>>>>

affiliations:
  - id: 1
    name: \textit{Computational Science Lab, Informatics Institute, University of Amsterdam, PO Box 94323, Amsterdam, 1090GH, the Netherlands}
  - id: 2
    name: \textit{Department of Public and Occupational Health, Amsterdam Public Health Research Institute, Amsterdam UMC, University of Amsterdam, Amsterdam, the Netherland}
  - id: 3
    name: \textit{Institute for Advanced Study, University of Amsterdam, Oude Turfmarkt 147, Amsterdam, 1012GC, the Netherland}   
format: 
  # docx:
  #   number-sections: true
  #   appendix: true
  pdf:
    pdf-engine: pdflatex
    fontfamily: palatino
    #toc: true
    number-sections: true
    appendix: true
    colorlinks: true
    keep-tex: true
    geometry: 
      #- showframe
      - top=40mm
      - left=45mm
      - right=45mm
      - bottom=40mm
    template-partials:
      - title.tex
    include-in-header: header.tex
    crossref:
      appendix-title: "App."
      appendix-delim: ":"
date: today
execute:
  echo: false
  warning: false
bibliography: references_draft.bib
nocite: |
  @*
csl: nature.csl
abstract: |
  \noindent This study investigates how the internal structure of depression–precariousness dynamics shapes responsiveness to external support. Using cross-sectional data from the HELIUS study, we first apply cycle-capable causal discovery algorithms to estimate directional dependencies between financial stress, domain-specific precariousness, and depressive symptoms. Our multi-resolution analysis—spanning both aggregate and symptom-level constructs—highlights financial stress as a plausible upstream driver of depression, with symptoms such as sleep disturbance and depressed mood emerging as initiators of broader social vulnerability.
  To translate these structural insights into dynamic predictions, we construct a family of analytically calibrated linear stochastic models. Simulations reveal that variations in internal structure can substantially alter system responses to financial stress reduction. Some models show rapid but transient improvements in depression, while others adjust more slowly yet sustain gains over time. Precariousness tends to respond more immediately across all configurations, though its trajectory also shifts subtly with structural variation.
  Together, these findings demonstrate that intervention outcomes depend not only on the magnitude of external input but also on latent properties of the system. More broadly, this study illustrates how combining causal discovery with dynamical modeling supports principled, simulation-based exploration of intervention scenarios—grounded in data, yet able to probe counterfactual structure. This approach offers a scalable framework for diagnosing hidden system constraints and informing targeted public health strategies in complex psychosocial domains.
---

@@

This study aims to investigate the causal mechanisms linking precariousness and depression and how those mechanisms shape system responses to external intervention. Specifically, we ask: *How does the internal configuration of the precariousness–depression relationship affect the outcomes of stress-reducing interventions?*
To address this question, we analyze cross-sectional data from more than 21,000 participants in the HELIUS (HEalthy LIfe in an Urban Setting) study [@stronks2013unravelling; @snijder2017cohort] using a two-step approach. First, we apply cycle-capable causal discovery algorithms to infer directional relationships among financial stress, domain-specific precariousness, and depressive symptoms. By examining both aggregated measures (a depression sum score and a composite precariousness index) and disaggregated indicators (individual symptoms and life-domain factors), we capture both broad patterns and fine-grained pathways. Second, we translate these insights into a computational dynamical model that simulates how depression and precariousness co-evolve under varying levels of external stress. This framework enables us to examine how system properties—such as feedback strength and stochastic noise—shape responsiveness to intervention.

@fig-concept presents this high-level model, derived from the causal discovery analysis and serving as the foundation for our simulation experiments. It highlights financial stress ($S_i$) as an upstream driver influencing both depressive symptoms ($D_i$) and social precariousness ($P_i$). Depression contributes to broader social vulnerability by feeding into precariousness, while precariousness feeds back into depression, creating reinforcing cycles. This framework illustrates that intervention effects depend not only on external inputs but also on the internal feedback architecture of the system. Full methodological details, including data processing, algorithmic specifications, and model calibration procedures, are provided in the [Methods](#sec-methods) section.

<!-- When the effects of depression symptoms onto the social precarity subsystem are weaker than the reverse path, there is higher sensitivity of depression to intervention, reflecting both a faster reduction post financial intervention but also quicker relapse once financial stresses are reestablished. -->
<!-- By integrating causal discovery with dynamic simulation, we demonstrate that intervention strategies can be tailored to the underlying architecture of vulnerability. -->

```{r}
#| label: fig-concept
#| fig-cap: "The model abstracts the main dependencies revealed by causal discovery into a compact system of dynamic interactions. Financial stress ($S_i$) is treated as an upstream driver that influences both depressive symptoms ($D_i(t)$) and social precariousness ($P_i(t)$). Depression and precariousness also influence each other through feedback loops, creating the potential for reinforcing cycles of vulnerability. Arrows represent influence coefficients ($\\alpha$), which correspond to terms in the system of stochastic differential equations used for simulation. This high-level structure served as the foundation for our simulation experiments. Full methodological details, refer to the [Methods](#sec-methods) section."
#| fig-align: center
#| out-width: 40%

@@

We present the results in two parts. First, we report the causal structures inferred from the HELIUS data at the levels of aggregate depression, individual symptoms, and aggregated precariousness. Second, we embed these findings in a simplified dynamical model to simulate how alternative feedback structures shape the system’s response to stress-reducing interventions.

@@

@@

```{r}
#| label: fig-intervention
#| fig-align: center
#| out-width: 80%
#| fig-cap: >
#|   System response to external intervention across variations in internal structure.
#|   ***Panel A**: Mean levels of depression ($D$) and precariousness ($P$) at time steps 100, 300, and 500, across models that differ in the coupling coefficient $\alpha_{DP}$, which represents the strength of depression’s influence on precariousness in the dynamical model. The x‑axis shows intervention strength, defined as the proportional reduction of financial stress ($S$) relative to its baseline (with higher values indicating more external financial support). Each line corresponds to a different model defined by a specific $\alpha_{DP}$ value, color-coded from low (purple) to high (yellow). Models with lower $\alpha_{DP}$ exhibit faster early reductions in $D$, while higher $\alpha_{DP}$ leads to more immediate declines in $P$.
#|   
#|   **Panel B**: Full time trajectories for two models—one with no coupling ($\alpha_{DP} = 0$) and one with strong coupling ($\alpha_{DP} = 0.65$)—illustrating system response and recovery following intervention onset (step 300) and withdrawal (step 1500).
#|   Depression responds more gradually and persistently in the high $\alpha_{DP}$ model, whereas precariousness adjusts more quickly.*

@@

# Methods {#sec-methods}

# Data Availibility
The data used in this study come from the HELIUS (HEalthy LIfe in an Urban Setting) cohort. Due to privacy and ethical restrictions, the raw HELIUS data are not publicly available. Researchers who wish to access the data can submit a request to the HELIUS study team via the study website (<https://heliusstudy.nl>), following the governance procedures of the HELIUS Executive Board.

# Code Availibility
All analysis scripts used for causal discovery, model calibration, and simulation are available at <https://github.com/KyuriP/Precarity_HELIUS>. The repository contains documented R code and instructions for reproducing the analyses reported in this manuscript.
